---
title: "06_Fst_analysis"
author: "Erica Robertson"
date: "2025-01-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="~/Desktop/GCRF/GCRF_Pub")
```

So we're going to run an Fst analysis using the program OutFLANK. This program takes a lot of data and so everything is going to be run on the cluster.

The first thing that needs to be done is splitting up the main .vcf file, ROFI.10x.pass-maf-0.05.SNP.biallelic.8miss.rmdup.rename.vcf.gz, into smaller files based on scaffolds. I used the comms file to split them into groups of roughly the same size. When there were two scaffolds in a group I separated their names with a comma. I did all this using grep functions in bbedit, so not reproducible, sorry. The first couple lines looked like this:

Scaffold_1__1_contigs__length_151072562
Scaffold_2__1_contigs__length_114774641
Scaffold_3__1_contigs__length_112408489
Scaffold_4__1_contigs__length_75720081
Scaffold_5__1_contigs__length_71959612
Scaffold_6__1_contigs__length_71876329
Scaffold_7__1_contigs__length_62267870
Scaffold_8__1_contigs__length_35685229
Scaffold_9__1_contigs__length_6533178
Scaffold_10__1_contigs__length_4747999
Scaffold_11__1_contigs__length_4700999
Scaffold_12__1_contigs__length_2025999,Scaffold_13__1_contigs__length_2412999
Scaffold_14__1_contigs__length_5954999

Then I wrote a script that uses bcftools to split up the main vcf by region based on the lines in scaffold_groups.txt.

```{bash}
#!/bin/bash
#SBATCH --job-name=subset
#SBATCH --output=subset.%j.out
#SBATCH --error=subset.%j.err
#SBATCH -t 1:00:00
#SBATCH -p amilan
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 5
#SBATCH --mem=12G
#SBATCH --mail-type=END,FAIL
#SBATCH  --mail-user=ericacnr@colostate.edu

source ~/.bashrc

set -x

conda activate bioinf

scaffold_file="/scratch/alpine/ericacnr@colostate.edu/GCRF_Pub/06.fst/scaffold_groups.txt"
input_vcf="/scratch/alpine/ericacnr@colostate.edu/GCRF_Pub/01.fastq_processing/ROFI.10x.pass-maf-0.05.SNP.biallelic.8miss.rmdup.rename.vcf.gz"

output_dir="/scratch/alpine/ericacnr@colostate.edu/GCRF_Pub/06.fst/scaffold_groups"

mkdir -p "$output_dir"

while IFS= read -r line; do
  # Define the output name based on the line number
  output_name="scaffold_group_${line_number}"

  # Subset the VCF for the current group of scaffolds
  bcftools view -r "$line" -Oz -o "$output_dir/${output_name}.vcf.gz" "$input_vcf"

  # Increment line counter
  ((line_number++))
done < "$scaffold_file"
```

Then, because of the way that OutFLANK want it's data organized for input, I ran vcftools 012 function on all the subset vcf files to turn the snps from 0/0, 0/1, 1/1 format to 0,1,2.

I ran this with a for loop that input's each vcf file from the scaffold_group directory. Have to be in the scaffold_group directory for it to work right.

for vcf in `ls *.vcf.gz`; do echo $vcf; sbatch ../02.vcftools_012.sbatch $vcf ;done

```{bash}
#!/bin/bash
#SBATCH --job-name=012
#SBATCH --output=012.%j.out
#SBATCH --error=012.%j.err
#SBATCH -t 1:00:00
#SBATCH -p amilan
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 5
#SBATCH --mem=12G
#SBATCH --mail-type=END,FAIL
#SBATCH  --mail-user=ericacnr@colostate.edu

source ~/.bashrc

set -x

conda activate bioinf

#for vcf in `ls *.vcf.gz`; do echo $vcf; sbatch ../02.vcftools_012.sbatch $vcf ;done

vcf=$1

vcftools --gzvcf $vcf --012 --out $vcf
```

This creates files called scaffold_group_0.vcf.gz.012, for example.

Now that those are ready, I'm going to run OutFLANK on every group separately. This is done with the script creatively named OutFLANK_script.R

This loops through every .vcf.gz.012 file in the directory and separately puts it into OutFLANK to get an fst output file for the group.
```{r}
library(vcfR)
library(OutFLANK)
library(data.table)
library(tidyverse)

files <- c(0,1:43)

for (n in files){
  vcf.012_path <- paste0("scaffold_groups/scaffold_group_",n,".vcf.gz.012")
  vcf.012 <- fread(vcf.012_path)
  #has missing data as a -1, need it to be a 9
  for (col in seq_along(vcf.012)) {
    set(vcf.012, i = which(vcf.012[[col]] == -1), j = col, value = 9)
  }
  unique(vcf.012$V2)
  vcf.012 <- vcf.012[,-1]

  #want the column names to be the snp names so the fst values are interpretable later
  snp.pos_path <- paste0("scaffold_groups/scaffold_group_",n,".vcf.gz.012.pos")
  snp.pos <- fread(snp.pos_path)
  snp.pos[, result := str_c(str_extract(V1, "Scaffold_[0-9]+"), V2, sep = ".")]
  snp.names <- snp.pos$result

  #changing the colnames and removing the individuals I don't want
  colnames(vcf.012) <- snp.names
  SNPmat <- vcf.012[c(-98,-154:-159,-20:-23, -146),]

  #the above would match the rs values, but we need the locusNames to match the SNPmat names, which are formatted differently
  locusNames <- colnames(SNPmat)

  #make a list that has the population data for all the individuals
  #I pulled the first column from a .fam file that has the pops as 1 or 2
  popNames <- readLines("/scratch/alpine/ericacnr@colostate.edu/GCRF_Pub/06.fst/pop_list.txt")

  #now using all this to make an OutFLANK formatted input file
  fst <- MakeDiploidFSTMat(SNPmat=SNPmat,locusNames=locusNames,popNames=popNames)
  file_name <- paste0("scaffold_groups/scaffold_group_"n,".vcf.gz.012_fst.csv")
  write.csv(fst, file_name)
}
```

This is run using the script 03.OUTFLANK_script.R.sbatch

```{bash, label="03.OUTFLANK_script.R.sbatch"}
#!/bin/bash
#SBATCH --job-name=OutFLANKtest
#SBATCH --output=OutFLANKtest.%j.out
#SBATCH --error=OutFLANKtest.%j.err
#SBATCH -t 3:00:00
#SBATCH -p amilan
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 30
#SBATCH --mem=100G
#SBATCH --mail-type=END,FAIL
#SBATCH  --mail-user=ericacnr@colostate.edu

source ~/.bashrc

set -x

conda activate R

Rscript OutFLANK_script.R
```

The output are then joined together and evaluated for outliers in the script processing_OutFLANK_output.R.

```{r, label="processing_OutFLANK_output.R"}
library(data.table)
library(tidyverse)
library(OutFLANK)

setwd("/scratch/alpine/ericacnr@colostate.edu/GCRF_Pub/06.fst/scaffold_groups")

csv_files <- list.files(pattern = "\\.csv$", full.names = TRUE)

data_list <- lapply(csv_files, read.csv)

names(data_list) <- basename(csv_files)

combined_data <- do.call(rbind, lapply(csv_files, read.csv))


outflank.output.01 <- OutFLANK(combined_data, LeftTrimFraction=0.05,
         RightTrimFraction=0.005, Hmin=0.2, 2,
         qthreshold=0.01)

png("../fst_plot_.01.png", width = 800, height = 600)  # Open a PNG device and specify size

fst.plot <- OutFLANKResultsPlotter(outflank.output.01,
                                   withOutliers = TRUE,
                                   NoCorr = TRUE,
                                   Hmin = 0.2,
                                   binwidth = 0.005,
                                   Zoom = TRUE,
                                   RightZoomFraction = 0.01,
                                   titletext = NULL)
dev.off()

str(outflank.output.01)

fst.results <- outflank.output.01$results
colnames(fst.results)[colnames(fst.results) == "LocusName"] <- "rs"
write.csv(fst.results, "../OutFLANK_fst_results.csv")
```

This outputs a plot of the fst values past the 0.1 threshold (to get a better look at the tail end of the snps) as well as a final OutFLANK_fst_results.csv to the 06.fst directory.

It's run using
```{bash, label="04.processing_OutFLANK_output.R.sbatch"}
#!/bin/bash
#SBATCH --job-name=OutFLANKproc
#SBATCH --output=OutFLANKproc.%j.out
#SBATCH --error=OutFLANKproc.%j.err
#SBATCH -t 3:00:00
#SBATCH -p amilan
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 30
#SBATCH --mem=100G
#SBATCH --mail-type=END,FAIL
#SBATCH  --mail-user=ericacnr@colostate.edu

source ~/.bashrc

set -x

conda activate R

Rscript processing_OutFLANK_output.R
```

List of 6
 $ FSTbar               : num 0.00449
 $ FSTNoCorrbar         : num 0.0141
 $ dfInferred           : num 2
 $ numberLowFstOutliers : int 0
 $ numberHighFstOutliers: int 831
 $ results              :'data.frame':  7739399 obs. of  16 variables:
  ..$ X               : int [1:7739399] 1 2 3 4 5 6 7 8 9 10 ...
  ..$ LocusName       : chr [1:7739399] "Scaffold_1.195" "Scaffold_1.213" "Scaffold_1.683" "Scaffold_1.751" ...
  ..$ He              : num [1:7739399] 0.12 0.183 0.133 0.184 0.117 ...
  ..$ FST             : num [1:7739399] -0.011958 -0.014773 -0.002232 0.000773 -0.008151 ...
  ..$ T1              : num [1:7739399] -7.16e-04 -1.35e-03 -1.49e-04 7.16e-05 -4.78e-04 ...
  ..$ T2              : num [1:7739399] 0.0599 0.0913 0.0666 0.0926 0.0587 ...
  ..$ FSTNoCorr       : num [1:7739399] 0.00155 0.00055 0.00885 0.01648 0.00333 ...
  ..$ T1NoCorr        : num [1:7739399] 9.31e-05 5.03e-05 5.90e-04 1.53e-03 1.95e-04 ...
  ..$ T2NoCorr        : num [1:7739399] 0.06 0.0914 0.0666 0.0927 0.0587 ...
  ..$ meanAlleleFreq  : num [1:7739399] 0.936 0.898 0.929 0.898 0.938 ...
  ..$ indexOrder      : int [1:7739399] 1 2 3 4 5 6 7 8 9 10 ...
  ..$ GoodH           : chr [1:7739399] "lowH" "lowH" "lowH" "lowH" ...
  ..$ qvalues         : num [1:7739399] NA NA NA NA NA ...
  ..$ pvalues         : num [1:7739399] NA NA NA NA NA ...
  ..$ pvaluesRightTail: num [1:7739399] NA NA NA NA NA ...
  ..$ OutlierFlag     : logi [1:7739399] FALSE FALSE FALSE FALSE FALSE FALSE ...
  
$FSTbar: Mean FST of loci not flagged as outliers, using the sample-size correction (recommend reporting in publication)
$FSTNoCorrbar: Mean FST of loci not flagged as outliers, without using the sample-size correction
$dfInferred: Inferred df for the chi-square distribution from the OutFLANK procedure (recommend reporting in publication).
$numberLowFstOutliers: Number of loci flagged as having a significantly low FST (not reliable)
$numberHighFstOutliers: Number of loci flagged as having a significantly high FST.
$results: a data frame with a row for each locus. This data frame includes all the original columns in the input data set, and six new ones:
$results$indexOrder: the original row number of the locus in the input data set
$results$GoodH: Boolean variable which is TRUE if the expected heterozygosity is greater than the Hemin set by the input to OutFLANK
$results$OutlierFlag: TRUE if the method identifies the locus as an outlier, FALSE otherwise, based on the q-threshold
$results$q: the q-value for the test of neutrality for the locus, based on the right-tailed P-value.
$results$pvalues: the two-tailed P-value for the test of neutrality for the locus. Not reliable, because OutFLANK does not predict the left-tailed P-values well.
$results$pvaluesRightTail: the one-sided (right tail) P-value for the locus



Then, we move that output over here to process and plot the results.

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/GCRF_Pub/06.fst/OutFLANK_fst_results_0.005.csv /Users/ericarobertson/Desktop/GCRF/GCRF_Pub/analysis/Fst/OutFLANK_output
```

You can try and do this on your own computer but mine really struggles... I'm going to do this in an interactive node on the cluster
```{r}
library(data.table)
library(dplyr)
fst.results <- fread("analysis/Fst/OutFLANK_output/OutFLANK_fst_results_0.005.csv")
fst.nona <- na.omit(fst.results$FST)
outliers <- fst.results %>% filter(OutlierFlag==TRUE) %>% select(rs, FST)
mean(outliers$FST)
min(outliers$FST)
max(outliers$FST)

write.csv(outliers, "results/fst_OutFLANK/OutFLANK_0.005FDR_outliers.csv", row.names = F)
```

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/GCRF_Pub/06.fst/OutFLANK_outliers.csv /Users/ericarobertson/Desktop/GCRF/GCRF_Pub/results/fst_OutFLANK
```

```{r}
fst_outliers <- read.csv("results/fst_OutFLANK/OutFLANK_0.005FDR_outliers.csv")
nrow(fst_outliers)

quantile_99 <- quantile(fst_outliers$FST, probs = 0.90, na.rm = TRUE)
print(quantile_99)

min(fst_outliers$FST)
mean(fst_outliers$FST)

fst_outliers.2 <- fst_outliers %>% filter(FST>= 0.2)
nrow(fst_outliers.2)
mean(fst_outliers.2$FST)

fst_outliers.3 <- fst_outliers %>% filter(FST>= 0.23)
nrow(fst_outliers.3)

write.csv(fst_outliers.3, "results/fst_OutFLANK/OutFLANK_outliers.23FST.csv", row.names = F)
```